Improving \textsl{Gaia} parallaxes with a data-driven model of the color--magnitude diagram

### 2017-03-08 paperDistance/paper.tex

We build a prior for the distance measurement of Gaia stars using the Gaia measured parallaxes plus APASS photometry to build a color-magnitude-like diagram that serves as the prior for distance measurements of Gaia stars. Instead of making assumptions about the physical distributions of stars in the galaxy, or using stellar models for the modeled HR diagram of stars, we build a prior from the data themselves.

@@

magSN = parallaxSN = 16 (excluding systematic parallax error in tgas)
train gaussian means, mus, Vs on progressively lower SN dataset

\begin{equation}
\label{eq:bayes}
P(\theta|\textbf{E}, \textbf{I}) = P(\theta ) \frac{P(\textbf{E} |\theta,\textbf{I})}{P(\textbf{E})},
\end{equation}

where $\theta$ is the model parameters that generate the evidence $\textbf{E}$, and P($\theta$) is the prior knowledge of the model, and $P(\textbf{E} |\theta,\textbf{I})$ is the likelihood of the evidence, given the model and some other prior assumptions. Our evidence is the parallaxes and G band magnitudes from Gaia, and the B and V band magnitudes from APASS. We build the prior from this evidence using extreme deconvolution

The data are a transformation of the dust corrected absolute magnitude, and the dust corrected colors.

\begin{equation}
\label{eq:data}
\begin{aligned}
\mathrm{data vector} \;\;\;\; &\vec{y} = [10^{0.2M_{J,n}}, J_n - K_n] \;  n \in \mathrm{stars} \\
\mathrm{with \; absolute \; magnitude} \;\;\;\; &M_{J,n} = J_n - DM_n - Q_JE(B-V)_n \\
& DM_n = -5\mathrm{log_{10}}\frac{1}{\varpi} \\
\mathrm{and \; color} \;\;\;\; &J_n - K_n = J_n - K_n - Q_{JK}E(B-V)_n

@@

We fit the data using XD which maximizes the likelihood

@@

Crossvalidation

@@

We use stars cross matched in the TGAS and APASS set of observations. We require there are no NaNs in any APASS photometric band, as well as every band having positive errors. We check that the match between APASS and TGAS is good using extreme APASS-WISE colors. We correct for dust using 3D Bayestar (Schlafly+Finkbeiner) with an estimate of the distance from sampling of posterior Gaia distances from Adrain. There is minimal dust to the matched stars. To build our model here we use the Gaia G band magnitude, parallax and APASS B-V color.

@@

Dust corrections are non trivial for stars with poor signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giants, which tend to have the lowest signal to noise, and can therefore have dust corrections $\sim 1-2$ magnitudes. To apply a dust correction to the 2MASS photometry, we first generate our prior using \xd\ on the observed, attenuated photometry. Using this raw prior we infer the distances to all the stars in our prior more precisely, increasing our signal to noise for each star and therefore minimize severe distances and their associated severe dust corrections. With more precise distance posteriors, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the Green et al 3D dust map at that distance. For each distance we take the median of the samples of the dust model. We apply each dust correction to our 2MASS photometry given by Equation \ref{eq:dust} where $K_{\lambda}$ = [3.626, 2.742, 3.303, 2.285, 1.698] for bands [B, V, g, r, i] respectively, taken from table (REF TABLE SFD), and $E(B-V)$ is the output of the Green et al dust models. We then regenerate our prior using the dust corrected photometry. We do not update the errors, but keep them the same as before the dust correction, which is wrong and under representing our errors but this leads again to a lighter deconvolution. Should I quote some statistics about the dust corrections? mostly small? maybe difference before after correcting for distance?

\begin{equation}
\label{eq:dust}
m_{\mathrm{band, corrected}} = m_{\mathrm{band}} - K_{\lambda} \times E(B-V)
\end{equation}

>>>>

@@

\begin{itemize}
\item The photometric uncertainties in the J band are small compared with the uncertainties in the parallax
\item Our empirical bayes approach is ok as long as we only infer information about the data contained within the prior
\item the uncertainties are gaussian
\item most stars lie where we expect them to on the CMD
\item number of gaussians that's efficient? or do x-validation
\item can't think
\end{itemize}

We would like to infer the distance to stars using the measured (noisy) parallax from Gaia. Gaia gives us the likelihood of the parallax of a star, given some unknown distance and uncertainties (assumed gaussian)
\begin{equation}
\label{eq:likelihood}
P(\varpi |d, \sigma^2_{\varpi}, \textbf{I})
\end{equation}

where $d$ is distance to the star, $\varpi$ is the observed parallax, and $\sigma^2_{\varpi}$ is the uncertainty in that measurement, and I are other various prior assumptions (mostly about the noise properties?). This is the probability of the data given the model, or distance, but we want to flip it around and infer the distance to the star given the measured parallax and it's uncertainty. We do this using probability theory/bayes rule, and turn the likelihood of the data into the posterior probablity of the distance

\begin{equation}
\label{eq:bayes}
P(d|\varpi, \sigma^2_{\varpi}, \textbf{I}) = \frac{P(\varpi |d, \sigma^2_{\varpi}, \textbf{I}) P(d)}{\textbf{Z}},
\end{equation}

where $P(d)$ is our prior knowledge of the distance to the star, and $\textbf{Z}$ is a normalization constant to make the posterior sum to 1. The strength in our inference method lies in how we generate our prior knowledge of the distances to stars. Here, instead of assuming the spatial distribution of stars in the milky way, or relying on stellar models, as previous work as shown, we build the color-magnitude diagram of the observed stars from their observations as our prior. The absolute magnitude of the CMD carries information about the distances to stars and where they can physically lie in this space. Toy model?

To generate our CMD prior we use the J and K band from 2MASS, as well as the parallax from Gaia. The data are a transformation of the dust corrected absolute magnitude, and the dust corrected colors.

\begin{equation}
\label{eq:data}
\begin{aligned}
\mathrm{data vector} \;\;\;\; &\vec{y} = [10^{0.2M_{J,n}}, (J- K)_n] \;  n \in \mathrm{stars} \\
\mathrm{with \; absolute \; magnitude} \;\;\;\; &M_{J,n} = J_n - DM_n - Q_JE(B-V)_n \\
& DM_n = -5\mathrm{log_{10}}\frac{1}{\varpi} \\
\mathrm{and \; color} \;\;\;\; &(J - K)_n = J_n - K_n - Q_{JK}E(B-V)_n

@@

where $E(B-V)$ is the output of the Schlafly dust model, and $Q$ is the correction factor for that band assuming some reddening law. To generate our prior, we fit the data in the CMD space using XD (Bovy et al) which deconvolves the data with its noise, which are heterogeneous measurement uncertainties. It generates the denoised, underlying distribution from which the noisey data were drawn from. It represents the distribution as a mixture of gaussians which maximizes the likelihood of the noisy data.

@@

Crossvalidation?

@@

We use stars crossmatched in TGAS, 2MASS and within the observing footprint of Panstarrs to access the Schlafly 3D dust model. We require the signal to noise of both the photometry and parallax be $> 0.001$, that the photometry have nonzero errors, and remove a small selection of 2MASS stars that have zero color and zero J band magnitude. We correct for dust using 3D Bayestar (Schlafly+Finkbeiner) with an estimate of the distance from iteratively sampling our posterior distances, more below.

@@

Dust corrections are non trivial for stars with poor signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giants, which tend to have the lowest signal to noise, and can therefore have dust corrections $\sim 1-2$ magnitudes. To apply a dust correction to the 2MASS photometry, we first generate our prior using \xd\ on the observed, attenuated photometry. Using this raw prior we infer the distances to all the stars in our prior more precisely, increasing our signal to noise for each star and therefore minimize severe distances and their associated severe dust corrections. With more precise distance posteriors, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the Green et al 3D dust map at that distance. For each distance we take the median of the samples of the dust model. We apply each dust correction to our 2MASS photometry given by Equation \ref{eq:dust} where $K_{\lambda}$ = [3.626, 2.742, 3.303, 2.285, 1.698, 0.709, 0.449, 0.302] for bands [B, V, g, r, i, J, H, K] respectively, taken from table (REF TABLE SFD), and $E(B-V)$ is the output of the Green et al dust models. We then regenerate our prior using the dust corrected photometry. We do not update the errors, but keep them the same as before the dust correction, which is wrong and under representing our errors but this leads again to a lighter deconvolution (Do we want to change this and add in the covariance for the dust?). Should I quote some statistics about the dust corrections? mostly small? maybe difference before after correcting for distance?

\begin{equation}
\label{eq:dust}
m_{\mathrm{band, corrected}} = m_{\mathrm{band}} - Q_{\lambda} \times E(B-V)
\end{equation}
